---
title: "RMS Titanic"
author: "Bhargavi Deshpande"
date: 01/27/2025
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission
<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**


```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look
<!-- -------------------------------------------------- -->

### __q1__ Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
df_titanic %>% glimpse
```

**Observations**:

- Class, Sex, Age, Survived, Number of persons in that classification of people

### __q2__ Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

- The Wikepedia page said that there were 2224 passengers and the data count says 2201
- Are there any differences?
  - Yes the Wikepedia page accoubt for approxiametely 23 souls
- If yes, what might account for those differences?
  - Since the dataset counts souls based on their classification, maybe there were individuals that did not fall under these classifications and therefore did not get counted
  - Another difference could be from how each source found their number of passengers. If the Wikepedia page determined their number passed on people boarded at the port while the data set had collected number of tickets, that could cause inconsistencies.

### __q3__ Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
d_survived <- df_titanic %>% filter( Survived == "Yes")
d_survived %>%
  ggplot() + geom_col(aes(x=Class, y=n, fill = Sex), position = "dodge")
```

**Observations**:

- Within the First and Second Class a lot more women survived the Titanic sinking. This makes sense from background knowledge, knowing that women and children who were in the upper classes were put onto lifeboats first. Additionally significantly more male crew members survived, however, this could either be because they were allowed to board lifeboats first or simply because at this time a majority of the crew were women not men. 

# Deeper Look
<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### __q4__ Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
d_survived <- df_prop %>% filter( Survived == "Yes")
d_survived %>%
  ggplot() + geom_col(aes(x=Class, y=Prop, fill = Sex), position = "dodge")
```





**Observations**:

- The graph demonstrates the proportion of people in each category that survived. This doesn't sway the visualization by only having the count of people who survived without knowing how many were on the ship in the first place. 
- Is there anything *fishy* going on in your plot?
Yes, there is something weird happening on the plot because it says all of the people, both men and women in the 1st and 2nd class survived. However, from the raw data we see that this is not the real case. I think this is because the data for the children is being mixed into the adult data since children are also classified with Male and Female in the data set. 
  

### __q5__ Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
d_survived <- df_prop %>% filter(Survived == "Yes", Age == "Adult")

d_survived %>%
  ggplot() + geom_col(mapping = aes(x = Class, y = Prop, fill = Sex), position = "dodge")

```


**Observations**:

- Now the proportion of women who survived is no longer 100%, it is more reflective of the data set. Compared to the graph that visualizes the total number of survived individuals in each category, this graph is much more insightful. This one allows us to see the pattern that women were able to leave the Titanic first compared to men. This can be seen through the crew data. While 150 more male crew members survived compared to female, 88% of female crew members survived compared to the 20% female.
- If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.
  - For q4, the code is plotting, of the people who survived, the proportion of the group (Female or Male) that falls under each class. However, the children are also categorized by Male or Female and are placed at the top of the data set. This means that as the code is plotting, the first grouping to meet the requirements of that bar on the graph are from the children data. For example, when looking at the Men who were in first class who survived, the first group of people who fall under this classification are the children in row 17. Since there were few children on board the Titanic, the ones in first and second class all survived. This is why the graph on q4 had the proportion for male and female being 1. Once the data was also filtered by age, like the graph of q5, the plot becomes more indicative of the proportions of adult males and females who survived in each class.

# Notes
<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
