---
title: "Aluminum Data"
author: "(Your name here)"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-data03-pivot-basics
---

*Purpose*: When designing structures such as bridges, boats, and planes, the design team needs data about *material properties*. Often when we engineers first learn about material properties through coursework, we talk about abstract ideas and look up values in tables without ever looking at the data that gave rise to published properties. In this challenge you'll study an aluminum alloy dataset: Studying these data will give you a better sense of the challenges underlying published material values.

In this challenge, you will load a real dataset, wrangle it into tidy form, and perform EDA to learn more about the data.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------------|------------------------|------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)
```

*Background*: In 1946, scientists at the Bureau of Standards tested a number of Aluminum plates to determine their [elasticity](https://en.wikipedia.org/wiki/Elastic_modulus) and [Poisson's ratio](https://en.wikipedia.org/wiki/Poisson%27s_ratio). These are key quantities used in the design of structural members, such as aircraft skin under [buckling loads](https://en.wikipedia.org/wiki/Buckling). These scientists tested plats of various thicknesses, and at different angles with respect to the [rolling](https://en.wikipedia.org/wiki/Rolling_(metalworking)) direction.

# Loading and Wrangle

<!-- -------------------------------------------------- -->

The `readr` package in the Tidyverse contains functions to load data form many sources. The `read_csv()` function will help us load the data for this challenge.

```{r load}
## NOTE: If you extracted all challenges to the same location,
## you shouldn't have to change this filename
filename <- "./data/stang.csv"

## Load the data
df_stang <- read_csv(filename)
df_stang
```

Note that these data are not tidy! The data in this form are convenient for reporting in a table, but are not ideal for analysis.

### **q1** Tidy `df_stang` to produce `df_stang_long`. You should have column names `thick, alloy, angle, E, nu`. Make sure the `angle` variable is of correct type. Filter out any invalid values.

*Hint*: You can reshape in one `pivot` using the `".value"` special value for `names_to`.

```{r q1-task}
## TASK: Tidy `df_stang`
df_stang_long <-
  df_stang %>%
    pivot_longer(
      names_to = c("var","angle"),
      names_sep = "_",
      values_to = "val",
      cols = starts_with("E") | starts_with("mu") | starts_with("nu"),
  ) %>%
  filter(val > 0) %>%
    pivot_wider(
      names_from = var,
      values_from = val,
    ) %>%
  mutate(angle = as.integer(angle)) %>%
  unnest(cols = c(E,nu))
df_stang
df_stang_long
```

Use the following tests to check your work.

```{r q1-tests}
## NOTE: No need to change this
## Names
assertthat::assert_that(
              setequal(
                df_stang_long %>% names,
                c("thick", "alloy", "angle", "E", "nu")
              )
            )
## Dimensions
assertthat::assert_that(all(dim(df_stang_long) == c(26, 5)))
## Type
assertthat::assert_that(
              (df_stang_long %>% pull(angle) %>% typeof()) == "integer"
            )

print("Very good!")
```

# EDA

<!-- -------------------------------------------------- -->

## Initial checks

<!-- ------------------------- -->

### **q2** Perform a basic EDA on the aluminum data *without visualization*. Use your analysis to answer the questions under *observations* below. In addition, add your own *specific* question that you'd like to answer about the data---you'll answer it below in q3.

```{r q2-task}
##
df_stang_sd_E <-
  df_stang_long %>%
  group_by(thick, angle, alloy) %>%
  summarize(sd_E = sd(E), .groups = "drop")
df_stang_sd_nu <-
   df_stang_long %>%
  group_by(thick, angle) %>%
  summarize(sd_nu = sd(nu), .groups = "drop")
df_stang_sds <-
  left_join(df_stang_sd_E, df_stang_sd_nu)
df_stang_sds
```

**Observations**:

-   Is there "one true value" for the material properties of Aluminum?
    -   No there is not "one true value" for the material properties of Aluminum
    -   The original data set shows two material property values for each thickness and angle of Aluminum, the modulus of elasticity (E) and Poisson's ratio (nu)
    -   However, for each thickness and angle there are multiple values for E and nu
    -   This Basic EDA found the standard deviation between the values of E and nu for each angle and thickness combination. There was only one form of Aluminum that had one "true" value for E as the standard deviation was 0. The rest of the E and nu values for each thickness and angle combination deviated from each other.
-   How many aluminum alloys are in this data set? How do you know?
    -   1 because all of the values under the "alloy" column are the same, al_24st
-   What angles were tested?
    -   0, 45, and 90
-   What thicknesses were tested?
    -   0.022, 0.032, 0.064, 0.081
-   Are the modulus of elasticity (E) or Poisson's ratio (nu) strongly related to the thickness of the aluminum

## Visualize

<!-- ------------------------- -->

### **q3** Create a visualization to investigate your question from q2 above. Can you find an answer to your question using the dataset? Would you need additional information to answer your question?

```{r q3-task}
df_stang_long %>%
  ggplot(aes(x = angle, y = E, color = as_factor(thick))) + 
  geom_point()
df_stang_long %>%
  ggplot(aes(x = angle, y = nu, color = as_factor(thick))) + 
  geom_point()
```

**Observations**:

-   From the data set given neither the modulus of elasticity or poisson's ratio are strongly correlated to thickness
-   Modulus of Elasticity v.s. Angle:
    -   It does seem that the thickest aluminum for all angles had the smallest modulus of elasticity. This would mean that aluminum at a thickness of 0.08 requires the least amount of force to deform it, which is interesting.
    -   Looking at higher thickness on modulus of elasticity graph, there isn't a strong trend that shows that the thinner or thicker a metal is the higher the modulus of elasticity is. For the angles 45 and 90, however, the values associated with the thinnest metals did have the highest modulus of elasticity.
    -   There are also fewer points on the graph shown because for each angle there were some modulus of elasticity values that were the same across thicknesses, so more data points existed but they got plotted over each other.
-   Poisson's Ratio v.s. Angle:
    -   There is less of a general trend in this graph
    -   Each angle value has a large spread of nu values
    -   Additionally, the thickness does not form a gradient on any angle nor is there a trend that shows that the thickness of the metal correlates to a higher or lower Poisson's Ratio
    -   The values associated with thicker aluminum are generally have a lower nu value but it is not consistent across all angles of aluminum
-   Having a lot more data would make these graphs more accurate and far more representative of the true behavior of aluminum
    -   It would be helpful to have data taken of angles within the ranges between 0, 45, and 90
    -   Also, having a larger spread of thicknesses
    -   Finally, have more data for each instance would better show common ranges, outliers, and more accurate generalization about each instance

### **q4** Consider the following statement:

> "A material's property (or material property) is an intensive property of some material, i.e. a physical property that does not depend on the amount of the material."[2]

Note that the "amount of material" would vary with the thickness of a tested plate. Does the following graph support or contradict the claim that "elasticity `E` is an intensive material property." Why or why not? Is this evidence *conclusive* one way or another? Why or why not?

```{r q4-vis}
## NOTE: No need to change; run this chunk
df_stang_long %>%
  ggplot(aes(nu, E, color = as_factor(thick))) +
  geom_point(size = 3) +
  theme_minimal()
```

**Observations**:

-   Does this graph support or contradict the claim above?
    -   The graph contradicts the claim above to an extent
    -   When the material has a thickness of 0.081, the material properties of nu and E are clustered together with no other thickness instances in that area. This would imply that thickness is a contributing factor to the material property
    -   As you move up the graph, the data points are less separated by thickness with the values associated with three thicknesses being clustered together.
    -   The thinnest material is generally higher up and together, but not by a significant amount
-   Is this evidence *conclusive* one way or another?
    -   I don't believe this evidence is conclusive one way or another because only one value of thickness is actually seen to have starkly different material properties than the others.
    -   This difference would indicate that thickness may play a role; however as thicknesses decrease there seems to be less of a trend in terms of material properties compared to thickness.
    -   There is also not enough data taken to make a claim like this as each thickness 6-8 data points associated with it and there are only four thicknesses being tested.

# References

<!-- -------------------------------------------------- -->

[1] Stang, Greenspan, and Newman, "Poisson's ratio of some structural alloys for large strains" (1946) Journal of Research of the National Bureau of Standards, (pdf link)[<https://nvlpubs.nist.gov/nistpubs/jres/37/jresv37n4p211_A1b.pdf>]

[2] Wikipedia, *List of material properties*, accessed 2020-06-26, (link)[<https://en.wikipedia.org/wiki/List_of_materials_properties>]
