---
title: "Gapminder"
author: "(Your name here)"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
editor_options: 
  markdown: 
    wrap: 72
---

*Purpose*: Learning to do EDA well takes practice! In this challenge
you'll further practice EDA by first completing a guided exploration,
then by conducting your own investigation. This challenge will also give
you a chance to use the wide variety of visual tools we've been
learning.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics
define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------------|------------------------|------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and
supporting files (`report_files/` folder) when you are done! Then submit
a link to Canvas. **Your Challenge submission is not complete without
all files uploaded to GitHub.**

```{r setup}
library(tidyverse)
library(gapminder)
library(ggrepel)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is
an independent organization that seeks to educate people about the state
of the world. They seek to counteract the worldview constructed by a
hype-driven media cycle, and promote a "fact-based worldview" by
focusing on data. The dataset we'll study in this challenge is from
Gapminder.

# Guided EDA

<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to
the high-level process we're going through---after this guided round
you'll be responsible for doing another cycle of EDA on your own!

### **q0** Perform your "first checks" on the dataset. What variables are in this dataset?

```{r q0-task}
gapminder 
```

**Observations**:

-   Country, Continent, Year, Life Expectancy, Population, GDP Per
    Capita

### **q1** Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble.
(Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- 
  gapminder %>%
  pull(year) %>%
  max()
  
year_min <- 
  gapminder %>%
  pull(year) %>%
  min()
year_max
year_min
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### **q2** Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document
your challenges and attempt to produce the most informative visual you
can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder %>%
  filter(
    year == year_min,
    gdpPercap < 90000,
  ) %>%
  ggplot + 
  geom_point(aes(x=continent, y = gdpPercap))
gapminder %>%
  filter(
    year == year_min,
  ) %>%
  ggplot + 
  geom_point(aes(x=continent, y = gdpPercap)) + scale_y_log10()
```

**Observations**: - For the first graph, an outlier from the dataset
that was in the continent of Asia was removed to allow for the rest of
the data set to be better spread and clear

-   The two countries in Oceania are close to the GDP of Europe
-   Apart from the outlier country in Asia, both the continents of
    Africa and Asia have lower GDPs per Capita which tracks with
    historical context of both of these continents
-   The Americas are grouped together, which is something I have not
    seen often. I think it would be interesting to see the spread when
    the Americas are separated between North America and South America.
    I also wonder how much the mean is being pulled by the upper points
    in the Americas versus the rest of the data points.

**Difficulties & Approaches**:

-   There was one country in Asia that was an outlier in the data set.
    It had a GDP Per Capita of above 90,000. Including this point in the
    graph made the rest of the data visualization cluster together at
    the bottom and no spread was identifiable
-   As a result I excluded data points, which was just the one point,
    greater than 90,000 in order to see the rest of the data more
    clearly.
-   Another approach I also looked at was adding a log scale on the
    y-axis. This pulled the "outlier" point in towards the rest of the
    dataset and also made the bottom portion of the data set a lot
    clearer.
-   I kept both visualizations in because I wanted to compare whether
    using a log axis or removing the outlier would show the separation
    between the points better without misrepresenting the data.

### **q3** You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}

whisk_gapminder <-
  gapminder %>%
  filter(year == year_min, gdpPercap < 90000 ) %>%
  group_by(continent) %>%
  summarize(
    low_whisk = boxplot.stats(gdpPercap)$stats[1],
    high_whisk = boxplot.stats(gdpPercap)$stats[5]
  )
  
whisk_gapminder


filt_gapminder <-
  gapminder %>%
  filter(year==year_min, gdpPercap < 90000) %>%
  left_join(whisk_gapminder, by = "continent") %>%
  mutate(in_range = gdpPercap >= low_whisk & gdpPercap <= high_whisk)

filt_gapminder



ggplot(
  filt_gapminder,
  aes(continent, gdpPercap)
) + geom_boxplot(outlier.shape = NA) +
  geom_jitter(data = filt_gapminder %>% 
    filter(in_range == FALSE),
    aes(color=country))


```

**Observations**:

-   There were ten total outliers based on this graph
-   I determined outliers based on using the box and whisker plot
    statistics. The values that were beyond the low whisker and upper
    whisker were considered outliers because they were relatively
    distant from the bulk distribution of the continent.
-   I filtered out the one largest outlier, that was greater than 90000,
    because it would make the rest of the point illegible as well as
    skew the boxplots -I think it's really interesting how the outliers
    are all above the high whisker and not below, I think this might be
    because of the proportionally large number of countries within a
    continent that are below the average GDP per capita of that
    continent and the very few countries who do have a high GDP per
    capita are significantly greater
-   Filtering this data does make me wonder how much taking an outlier
    away from the data changes the accuracy of the box and whisker plots
    or any other statistical analysis being done.

*Hint*: For the next task, it's helpful to know a ggplot trick we'll
learn in an upcoming exercise: You can use the `data` argument inside
any `geom_*` to modify the data that will be plotted *by that geom
only*. For instance, you can use this trick to filter a set of points to
label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### **q4** Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple
variables; think about using different aesthetics or facets.

```{r q4-task}
whiskmin_gapminder <-
  gapminder %>%
  filter(year == year_min, gdpPercap < 90000 ) %>%
  group_by(continent) %>%
  summarize(
    low_min_whisk = boxplot.stats(gdpPercap)$stats[1],
    high_min_whisk = boxplot.stats(gdpPercap)$stats[5]
  )
  


whiskmax_gapminder <-
  gapminder %>%
  filter(year == year_max, gdpPercap < 90000 ) %>%
  group_by(continent) %>%
  summarize(
    low_max_whisk = boxplot.stats(gdpPercap)$stats[1],
    high_max_whisk = boxplot.stats(gdpPercap)$stats[5]
  )
  


all_whisk <-
  whiskmin_gapminder %>%
  left_join(whiskmax_gapminder, by = "continent")


df<-
  gapminder %>%
  filter(year %in% c(1952, 2007)) %>%
  filter(gdpPercap < 90000) %>%
  left_join(all_whisk, by = "continent") %>%
  
  mutate(
    min_in_range = case_when(
      year == 1952 ~ gdpPercap >= low_min_whisk & gdpPercap <= high_min_whisk,
    TRUE ~ NA),
    max_in_range = case_when(
      year == 2007 ~ gdpPercap >= low_max_whisk & gdpPercap <= high_max_whisk,
   TRUE ~ NA)
        )

ggplot(df, aes(continent, gdpPercap)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(data = df %>%
    filter((year == 1952 & min_in_range == FALSE) | (year == 2007 & max_in_range == FALSE)),
    aes(color=country)) +
    facet_wrap(~year)


```

**Observations**:

\- I think it is really interesting how in 2007 the whiskers for the box
and whisker plot were significantly longer in Asia and Europe than in
1952. This may be a result of a general growth and larger spread of GDP
per capita for countries in the 2000s compared to 1952.

\- Much like the previous graph, there are no outliers below the lower
whiskers for any year or country - It's also interesting and only four
countries remained outliers, above the GDP per Capita of the continent,
in 1952 and 2007. These countries were Gabon, South Africa, United
States, and Canada

\- While all of the outliers shown are outliers amongst their continent,
it's interesting to see how far apart the outliers are from each other.
Gabon is an oulier with a higher GDP of the rest of Africa with a GDP
per Capita of a little under 9000 while Canada is the lowest outlier in
the Americas with a GDP per Capita of 35000 in 2007.

# Your Own EDA

<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP
per capita at two time points. You can continue looking at outliers,
consider different years, repeat the exercise with `lifeExp`, consider
the relationship between variables, or something else entirely.

### **q5** Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
## 
filtnine_gapminder <-
  gapminder %>%
  filter(year == year_min, continent == "Asia", gdpPercap < 90000 ) %>%
  group_by(continent) %>%
  summarize(
    Q1 = boxplot.stats(gdpPercap)$stats[1],
    Q2 = boxplot.stats(gdpPercap)$stats[2],
    Q3 = boxplot.stats(gdpPercap)$stats[3],
    Lower_Whisker = boxplot.stats(gdpPercap)$stats[4],
    Upper_Whisker = boxplot.stats(gdpPercap)$stats[5],
    .groups = "drop") %>%
    mutate(Type = "Filtered")
    



nofilt_gapminder <-
  gapminder %>%
  filter(year == year_min, continent == "Asia") %>%
  group_by(continent) %>%
  summarize(
    Q1 = boxplot.stats(gdpPercap)$stats[1],
    Q2 = boxplot.stats(gdpPercap)$stats[2],
    Q3 = boxplot.stats(gdpPercap)$stats[3],
    Lower_Whisker = boxplot.stats(gdpPercap)$stats[4],
    Upper_Whisker = boxplot.stats(gdpPercap)$stats[5],
    .groups = "drop")%>%

    mutate(Type = "Unfiltered")
    

both_gapminder <-
  bind_rows(filtnine_gapminder, nofilt_gapminder)

gap_minder_long <-
both_gapminder %>%
  pivot_longer(
    cols = c(Q1, Q2, Q3, Lower_Whisker, Upper_Whisker),
    names_to = "Statistic",
    values_to = "GDP_Values"
  )
gap_minder_long

bar_plot <-
  ggplot(gap_minder_long, aes(x = Statistic, y = GDP_Values, fill = Type)) +
  geom_col(position = "dodge")
bar_plot

```

-   How much do the box and whisker plot values change when you do or
    don't filter out the outlier?
-   It is interesting that the Q1 values across both are identical, and
    the Q2 and Q3 values are not that different from each other but the
    whiskers are significantly different
-   I believe this is because there's is only one point being filtered
    out, so for calculating the quartiles, the quantities are not skewed
    much
-   The whisker values for the data without the outlier point removed
    were both greater because that outlier is pulling the values up.

```{r q5-task2}
## TASK: Your second graph
us_gdp <-
  gapminder %>%
  filter(country == "United States")
us_gdp
ggplot(us_gdp, aes(year, gdpPercap))+
  geom_point()
  
```

-   How has the GDP per Capita of the United States changed over time
    and can it be mapped to historical events?
-   It seems like there has been a linear growth of the GDP Per Capita
    of the United States from the years 1950 to 2007
-   Having more data would make this plot much more informative. Since
    there is only one data point every 5 years, the trend between years
    gets lost. However, for this case it may not affect it as much since
    I don't believe a country's GDP per Capita can change that
    significantly within five years
-   Additionally, have more recent data, even just 2008, would be really
    interesting to see if the 2008 recession had a significant impact on
    the country's GDP
-   In 1982 we see a data point that doesn't follow the linear trend.
    After doing some extra research I discovered that there was a
    significant recession, the worst since the Great Depression, from
    July 1981 to November 1982. This was caused by tight monetary
    regulation that was in an effort to fight the rising inflation.
-   It would be interesting to see if there were global economic
    declines during this time period by mapping the world's economically
    leading countries and seeing if there is a year where they all drop.

```{r q5-task3}
f_gap <-
  gapminder %>%
  filter(gdpPercap < 50000)

  ggplot(f_gap, aes(gdpPercap,lifeExp, color = continent))+
  geom_point() +
  facet_wrap(~year)
```

-   Are the GDP and life expectancy correlated in any way?

-   For each year, it does not seem like there is a trend between the
    GDP per capita of a country and their life expectancy

-   It seems more like when the country's GDP per capita is below 50000,
    there is a larger spread of life expentance of the country. However,
    as the country's GDP per Capita hits 10,000, the life expectancy is
    generally above 60

-   This trend is more clearly seen in the more recent years as the GDP
    per Capita has increased. As the GDP per capita gets higher, the
    spread of the life expectancy values decreases and there are no
    countries whose life expectancy is below 70

-   I think it's interesting how over time the locations of the coloring
    of each continent has not changed in relation to each other. Africa
    stays towards the bottom left in the span of 55 years while the
    European countries remain towards the top in life expectancy and GDP
    per Capita.
