---
title: "Massachusetts Highway Stops"
author: "(Your name here)"
date: 2020-
output:
  github_document:
    toc: true
---

*Purpose*: In this last challenge we'll focus on using logistic regression to study a large, complicated dataset. Interpreting the results of a model can be challenging---both in terms of the statistics and the real-world reasoning---so we'll get some practice in this challenge.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----|----|----|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

*Background*: We'll study data from the [Stanford Open Policing Project](https://openpolicing.stanford.edu/data/), specifically their dataset on Massachusetts State Patrol police stops.

```{r setup}
library(tidyverse)
library(broom)
```

# Setup

<!-- -------------------------------------------------- -->

### **q1** Go to the [Stanford Open Policing Project](https://openpolicing.stanford.edu/data/) page and download the Massachusetts State Police records in `Rds` format. Move the data to your `data` folder and match the `filename` to load the data.

*Note*: An `Rds` file is an R-specific file format. The function `readRDS` will read these files.

```{r q1-task}
## TODO: Download the data, move to your data folder, and load it
filename <- "C:\\Users\\bdeshpande\\Documents\\Data Science\\data-science-s2025\\challenges\\data\\MAstatepatrol.rds"
df_data <- readRDS(filename)

df_data
```

# EDA

<!-- -------------------------------------------------- -->

### **q2** Do your "first checks" on the dataset. What are the basic facts about this dataset?

**Observations**:

-   It has 24 columns
-   The columns describe the type of arrest, along multiple categories, and the person that was arrested
-   There are over 3 million rows of people arrested

Note that we have both a `subject_race` and `race_Raw` column. There are a few possibilities as to what `race_Raw` represents:

-   `race_Raw` could be the race of the police officer in the stop
-   `race_Raw` could be an unprocessed version of `subject_race`

Let's try to distinguish between these two possibilities.

### **q3** Check the set of factor levels for `subject_race` and `raw_Race`. What do you note about overlap / difference between the two sets?unq

```{r q3-task}
## TODO: Determine the factor levels for subject_race and raw_Race
unq_sub <-
unique(df_data$subject_race)
unq_raw <-
unique(df_data$raw_Race)

unq_raw
uniq_tib <-
  tibble(
    subject_race = unq_sub
  )
uniq_tib <- 
  bind_rows(
    uniq_tib, 
    tibble(subject_race = c(NA, NA)))
unc_tib <-
  bind_cols(
    uniq_tib,
    tibble(raw_Race = unq_raw)
  )

unc_tib
```

**Observations**:

-   What are the unique values for `subject_race`?
    -   white, hispanic, black, asian/pacific islander other, unknown
-   What are the unique values for `raw_Race`?
    -   White, Hispanic, Black, Asian or Pacific Islander, Middle Eastern or East Indian (South Asia), None-for no operator present citations only, A
-   What is the overlap between the two sets?
    -   White, Hispanic, Black, Asian/Pacific Islander
-   What is the difference between the two sets?
    -   The raw race is more specific and does not include an other option
    -   There is an option in the raw race for "None" presumably because the person was not there and not because they don't have a race.
    -   I don't know what "A" means
    -   Subject Race has two less rows

### **q4** Check whether `subject_race` and `raw_Race` match for a large fraction of cases. Which of the two hypotheses above is most likely, based on your results?

*Note*: Just to be clear, I'm *not* asking you to do a *statistical* hypothesis test.

```{r q4-task}
## TODO: Devise your own way to test the hypothesis posed above.
df_races <-
  df_data %>%
    mutate(
      sub_lower = str_to_lower(subject_race),
      raw_lower = str_to_lower(raw_Race)
    ) %>%
  select("raw_row_number", "sub_lower", "raw_lower")
df_boolean <-
  df_races %>%
    reframe(
      same = sub_lower == raw_lower
    ) %>%
  count(same)
df_boolean
  
df_races
```

**Observations**

Between the two hypotheses:

-   `race_Raw` could be the race of the police officer in the stop
-   `race_Raw` could be an unprocessed version of `subject_race`

which is most plausible, based on your results?

-   raw_Race is probable an unprocessed version of subject_race
-   It is unlikely that raw_Race is the race of the police officer in the stop because about 3.2 million cases out of 3.4 million cases had the same raw_Race and subject_race which seems a lot less possible than if the raw_race was an unprocessed version of subject_race.

## Vis

<!-- ------------------------- -->

### **q5** Compare the *arrest rate*---the fraction of total cases in which the subject was arrested---across different factors. Create as many visuals (or tables) as you need, but make sure to check the trends across all of the `subject` variables. Answer the questions under *observations* below.

(Note: Create as many chunks and visuals as you need)

```{r}
df_age <-
  df_data %>%
  group_by(
    subject_age,
    arrest_made
  ) %>%
  summarize(
    n=n()
  ) %>% 
  group_by(
    subject_age
  ) %>%
  mutate(
    total = sum(n),
    prop = n/total
  ) %>% 
  filter(
    arrest_made == TRUE
  )

df_age
df_age %>%
  ggplot(aes(subject_age, prop)) +
  geom_point() 
```

```{r}
df_sex <-
  df_data %>%
  group_by(
    subject_sex,
    arrest_made
  ) %>%
  summarize(
    n=n()
  ) %>% 
  group_by(
    subject_sex
  ) %>% 
  mutate(
    total = sum(n),
    prop = n/total
  ) %>% 
  filter(
    arrest_made == TRUE
  )
  
df_sex
df_sex %>% 
  ggplot() +
  geom_col(aes(x=subject_sex,y=prop, fill = arrest_made))
```

```{r}
df_grouprace <-
  df_data %>%
    group_by(
      subject_race,
      arrest_made
    ) %>%
    summarize(
      n=n()
    ) %>% 
    group_by(
      subject_race
    ) %>% 
    mutate(
      total = sum(n),
      prop = n/total
    ) %>% 
    filter(
      arrest_made == TRUE
    )

df_grouprace %>% 
  ggplot(aes(subject_race, prop, color = arrest_made)) +
  geom_point()
```

```         
```

**Observations**:

-   How does `arrest_rate` tend to vary with `subject_age`?
    -   Arrest rate tends to increase until the subject's age is in the mid-twenties, then the arrest rate generally decreases
-   How does `arrest_rate` tend to vary with `subject_sex`?
    -   Males have about double the arrest rate of women
    -   The difference in proportion is between 0.033 and 0.015
-   How does `arrest_rate` tend to vary with `subject_race`?
    -   When the subject's race is Hispanic, the arrest rate is the highest at 0.06
    -   The second highest arrest rate at 0.035 is when the subject's race is Black
    -   The lowest arrest rate is when the subject's race is unknown and the second lowest is Asian/Pacific Island
    -   I think the lowest arrest rate when the subject's race is unknown could be due to the fact that there are simply fewer people with that were not identified with a race.

# Modeling

<!-- -------------------------------------------------- -->

We're going to use a model to study the relationship between `subject` factors and arrest rate, but first we need to understand a bit more about *dummy variables*

### **q6** Run the following code and interpret the regression coefficients. Answer the the questions under *observations* below.

```{r q6-task}
## NOTE: No need to edit; inspect the estimated model terms.
fit_q6 <-
  glm(
    formula = arrest_made ~ subject_age + subject_race + subject_sex,
    data = df_data %>%
      filter(
        !is.na(arrest_made),
        subject_race %in% c("white", "black", "hispanic")
      ),
    family = "binomial"
  )

fit_q6 %>% tidy()

```

**Observations**:

-   Which `subject_race` levels are included in fitting the model?
    -   Hispanic, White
-   Which `subject_race` levels have terms in the model?
    -   Hispanic and White

You should find that each factor in the model has a level *missing* in its set of terms. This is because R represents factors against a *reference level*: The model treats one factor level as "default", and each factor model term represents a change from that "default" behavior. For instance, the model above treats `subject_sex==male` as the reference level, so the `subject_sexfemale` term represents the *change in probability* of arrest due to a person being female (rather than male).

The this reference level approach to coding factors is necessary for [technical reasons](https://www.andrew.cmu.edu/user/achoulde/94842/lectures/lecture10/lecture10-94842.html#why-is-one-of-the-levels-missing-in-the-regression), but it complicates interpreting the model results. For instance; if we want to compare two levels, neither of which are the reference level, we have to consider the difference in their model coefficients. But if we want to compare all levels against one "baseline" level, then we can relevel the data to facilitate this comparison.

By default `glm` uses the first factor level present as the reference level. Therefore we can use `mutate(factor = fct_relevel(factor, "desired_level"))` to set our `"desired_level"` as the reference factor.

### **q7** Re-fit the logistic regression from q6 setting `"white"` as the reference level for `subject_race`. Interpret the the model terms and answer the questions below.

```{r q7-task}
## TODO: Re-fit the logistic regression, but set "white" as the reference
## level for subject_race
fit_q7 <-
  glm(
    formula = arrest_made ~ subject_age + subject_race + subject_sex,
    data = df_data %>%
      mutate(
          subject_race = fct_relevel(subject_race, "white")
      ) %>%
      filter(
        !is.na(arrest_made),
        subject_race %in% c("white", "black", "hispanic")
      ),
    family = "binomial"
  )

fit_q7 %>% tidy()
```

**Observations**:

-   Which `subject_race` level has the highest probability of being arrested, according to this model? Which has the lowest probability?
    -   Hispanic
-   What could explain this difference in probabilities of arrest across race? List **multiple** possibilities.
    -   Systemic Racism
    -   Biases
    -   Maybe age
    -   Less generational wealth in the U.S. among families who immigrated leads to living in lower income communities that historically been targeted as areas with higher crime rates
-   Look at the set of variables in the dataset; do any of the columns relate to a potential explanation you listed?
    -   Potentially age
    -   County
    -   Whether a citation was given or not

One way we can explain differential arrest rates is to include some measure indicating the presence of an arrestable offense. We'll do this in a particular way in the next task.

### **q8** Re-fit the model using a factor indicating the presence of contraband in the subject's vehicle. Answer the questions under *observations* below.

```{r q8-task}
## TODO: Repeat the modeling above, but control for whether contraband was found
## during the police stop
fit_q8 <-
  glm(
    formula = arrest_made ~ subject_age + subject_race + subject_sex + contraband_found,
    data = df_data %>%
      mutate(
          subject_race = fct_relevel(subject_race, "white")
      ) %>%
      filter(
        !is.na(arrest_made),
        subject_race %in% c("white", "black", "hispanic")
      ),
    family = "binomial"
  )
fit_q8 %>% tidy()
```

**Observations**:

-   How does controlling for found contraband affect the `subject_race` terms in the model?
    -   It decreases the estimate for both Black and Hispanic races
    -   The subject's race being black has a lower estimate than when the subject is white
-   What does the *finding of contraband* tell us about the stop? What does it *not* tell us about the stop?
    -   It tells us that there is a lower proportion of Hispanic individuals and Black individuals that got arrested when contraband was found versus when contraband was not found
    -   It does not tell us anything about the baseline we are comparing these numbers to, when the subject is white
    -   It also does not tell us if the criteria for arrest changes when contraband is included in the case.

### **q9** Go deeper: Pose at least one more question about the data and fit at least one more model in support of answering that question.

```{r}
df_outcomes <-  
  df_data %>%
    group_by(
      subject_race,
      outcome
    ) %>%
    summarize(
      n=n()
    ) %>% 
    group_by(
      subject_race
    ) %>% 
    mutate(
      total = sum(n),
      prop = n/total
    ) %>% 
    filter(
      !is.na(outcome)
      )

df_outcomes %>% 
  ggplot(aes(subject_race, prop)) +
  geom_point()+
  facet_wrap(~ outcome) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

df_outcomes %>% 
  ggplot(aes(subject_race, prop, color = outcome)) +
  geom_point()+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
  
```

**Observations**:

-   Hispanics have the highest proportion of arrests, however, white people have the highest proportion of warnings given
-   White people have the lowest proportion of citations given which is interesting because a citation is the first, or lowest, punishment a person could get when stopped
-   When the race is unknown, the proportion of citations was really high, which could just be a coincident but is kind of interesting.

## Further Reading

<!-- -------------------------------------------------- -->

-   Stanford Open Policing Project [findings](https://openpolicing.stanford.edu/findings/).
